### 2016-10-26 sec-model.tex
Add a few more citations, minor cleanups.

\subsection{Formal definition of TL-HMM}

@@

\subsection{Parameter Estimation}
To learn the parameters of our model, we may attempt to use maximum
likelihood estimation. Unfortunately, a closed-form solution does not
exist, so we must appeal to the EM algorithm~\cite{Dempster:1977:JRSS}. In
particular, we propose a minor modification of the Baum-Welch
algorithm~\cite{Rabiner:1990:RSR} which is used to learn the parameters for
hidden Markov models. In the following sections, we will provide a brief
description of the original Baum-Welch algorithm, and then describe our
modification.

@@

Fortunately, the recursions for the $\alpha$ and $\beta$ variables remain
the same, as do the definitions of $\gamma_t(i)$ and $\xi_t(i,j)$ in the
E-step. The only change is in the updating equations in the M-step,
where we replace the update for $b_i(v)$ by a pair of updates: one for
$\pi^{(i)}_a$ one for $A^{(i)}_{ab}$. We denote the $m$-th action in the
$t$-th sequence for the $k$-th student as $o_{k,t,m}$. The two updates can
be written as
\begin{align}
  \pi^{(i)}_{a}
  &= \frac{\sum_{k=1}^M \sum_{t=1,a_1 = a}^T \gamma^{(\mathbf{o}_k)}_t(i)}
  {\sum_{k=1}^M \sum_{t=1}^T \gamma^{(\mathbf{o}_k)}_t(i)}, \text{ and }\\
  A^{(i)}_{ab}
  &= \frac{\sum_{k=1}^M \sum_{t=1}^T
  \sum_{m=2,a_{m-1}=a \land a_m = b}^{|\mathbf{o}_{k,t}|}
  \gamma^{(\mathbf{o}_k)}_t(i)}
  {\sum_{k=1}^M \sum_{t=1}^T \sum_{m=2}^{|\mathbf{o}_{k,t}|}
  \gamma^{(\mathbf{o}_k)}_t(i)}
\end{align}
where $a_m = o_{t,k,m}$ (for notational conciseness). Our modified EM
algorithm for HMMs with sequence observations is provided as part of the
\meta/ toolkit~\citemeta{}.

>>>>

\subsection{Formal Definition of TL-HMM}

@@

\subsection{Parameter Estimation}
To learn the parameters of our model, we may attempt to use maximum
likelihood estimation. Unfortunately, a closed-form solution does not
exist, so we must appeal to the EM algorithm~\cite{Dempster:1977:JRSS}. In
particular, we propose a minor modification of the Baum-Welch
algorithm~\cite{Rabiner:1990:RSR} which is used to learn the parameters for
hidden Markov models in an unsupervised setting. In the following sections,
we will provide a brief description of the original Baum-Welch algorithm,
and then describe our modification.

@@

Fortunately, the recursions for the $\alpha$ and $\beta$ variables remain
the same, as do the definitions of $\gamma_t(i)$ and $\xi_t(i,j)$ in the
E-step. The only change is in the updating equations in the M-step,
where we replace the update for $b_i(v)$ by a pair of updates: one for
$\pi^{(i)}_a$ one for $A^{(i)}_{ab}$. We denote the $m$-th action in the
$t$-th sequence for the $k$-th student as $o_{k,t,m}$. The two updates can
be written as
\begin{align}
  \pi^{(i)}_{a}
  &= \frac{\sum_{k=1}^M \sum_{t=1,a_1 = a}^T \gamma^{(\mathbf{o}_k)}_t(i)}
  {\sum_{k=1}^M \sum_{t=1}^T \gamma^{(\mathbf{o}_k)}_t(i)}, \text{ and }\\
  A^{(i)}_{ab}
  &= \frac{\sum_{k=1}^M \sum_{t=1}^T
  \sum_{m=2,a_{m-1}=a \land a_m = b}^{|\mathbf{o}_{k,t}|}
  \gamma^{(\mathbf{o}_k)}_t(i)}
  {\sum_{k=1}^M \sum_{t=1}^T \sum_{m=2}^{|\mathbf{o}_{k,t}|}
  \gamma^{(\mathbf{o}_k)}_t(i)}
\end{align}
where $a_m = o_{t,k,m}$ (for notational conciseness). Our modified EM
algorithm for TL-HMMs is provided as part of the \meta/
toolkit~\citemeta{}.

### 2016-10-26 sec-related-work.tex
Add a few more citations, minor cleanups.

Our model is based heavily upon the prior art of Hidden Markov
Models~\cite{Rabiner:1990:RSR} for sequence labeling tasks. These types of
models are widely applicable and have been used for tasks such as speech
recognition~\cite{Huang:1990:HMM} and part-of-speech
tagging~\cite{Jurafsky:2009:SLP}. A major challenge in applying HMMs successful to solve a problem is to design an appropriate architecture of the model, which always varies according to specific applications.

>>>>

Our model is based heavily upon the prior art of Hidden Markov
Models~\cite{Rabiner:1990:RSR} for sequence labeling tasks. These types of
models are widely applicable and have been used for tasks such as speech
recognition~\cite{Huang:1990:HMM} and part-of-speech
tagging~\cite{Jurafsky:2009:SLP}, and are a member of the more general
family of probabilistic graphical models~\cite{Koller:2009:PGM}.
A major challenge in applying HMMs and other graphical models successfully to
solve a problem is to design an appropriate architecture of the model, which
always varies according to specific applications.

### 2016-10-26 sec-results.tex
Add a few more citations, minor cleanups.

\section{Experiment Results}
As an analysis tool, the TL-HMM model provides with us the following two
patterns to characterize student behavior:
\begin{enumerate*}[label=(\arabic*)]
  \item latent state representations, and
  \item latent state transitions.
\end{enumerate*}
Thus to evaluate the proposed model, we conduct experiments to
qualitatively analyze both types of patterns discovered from empirical MOOC
log data.

@@

\begin{figure*}
  \centering
  \begin{tabular}{cccc}
    \textbf{State 0} & \textbf{State 1} & \textbf{State 2} & \textbf{State
    3}\\\hline
    \includegraphics[width=0.22\textwidth]{figures/text-2state/state0.png}
    &
    \includegraphics[width=0.22\textwidth]{figures/text-2state/state1.png}\\
    \includegraphics[width=0.22\textwidth]{figures/text-3state/state0.png}
    &
    \includegraphics[width=0.22\textwidth]{figures/text-3state/state1.png}
    &
    \includegraphics[width=0.22\textwidth]{figures/text-3state/state2.png}\\
    \includegraphics[width=0.22\textwidth]{figures/text-4state/state0.png}
    &
    \includegraphics[width=0.22\textwidth]{figures/text-4state/state1.png}
    &
    \includegraphics[width=0.22\textwidth]{figures/text-4state/state2.png}
    &
    \includegraphics[width=0.22\textwidth]{figures/text-4state/state3.png}
  \end{tabular}
  \caption{The evolution of states for increasing $K$ for the
  \protect\textretrieval{} MOOC.} % I don't know why \protect is needed
  \label{fig:text-state-evolution}
\end{figure*}

@@

\begin{figure*}
  \centering
  \begin{tabular}{cccc}
    \textbf{State 0} & \textbf{State 1} & \textbf{State 2} & \textbf{State
    3}\\\hline
    \includegraphics[width=0.22\textwidth]{figures/sustain-2state/state0.png}
    &
    \includegraphics[width=0.22\textwidth]{figures/sustain-2state/state1.png}\\
    \includegraphics[width=0.22\textwidth]{figures/sustain-3state/state0.png}
    &
    \includegraphics[width=0.22\textwidth]{figures/sustain-3state/state1.png}
    &
    \includegraphics[width=0.22\textwidth]{figures/sustain-3state/state2.png}\\
    \includegraphics[width=0.22\textwidth]{figures/sustain-4state/state0.png}
    &
    \includegraphics[width=0.22\textwidth]{figures/sustain-4state/state1.png}
    &
    \includegraphics[width=0.22\textwidth]{figures/sustain-4state/state2.png}
    &
    \includegraphics[width=0.22\textwidth]{figures/sustain-4state/state3.png}
  \end{tabular}
  \caption{The evolution of states for increasing $K$ for the
  \protect\sustain{} MOOC.} % I don't know why \protect is needed
  \label{fig:sustain-state-evolution}
\end{figure*}

We observe similar behavior in Figure~\ref{fig:sustain-state-evolution}
as we increase $K$ when fitting our TL-HMM on the \sustain{} MOOC. Again,
in the transition between $K=2$ and $K=3$ we discover forum behavior
patterns as a latent state. However, in the transition between $K=3$ to
$K=4$ we instead see a refining of that discovered forum state, where
state 2 captures asymmetric transition probabilities between ``forum thread
view'' and ``forum thread list''. These states could be seen as redundant,
in which case a more appropriate setting of $K=3$ may be appropriate for
the \sustain{} dataset.

@@

Thus, we might expect to see students that perform well in the course
preferring states 1 and 2 over states 0 and 3. To verify this, we took the
model we learned on the full training data and retrofit it to training data
consisting only of sequences produced by students in \textretrieval{} that
had perfect marks. To prevent the latent state meanings from drifting, we
forced the model parameters associated with their Markov model
representations to be fixed, in effect only learning initial and transition
probabilities for the top layer of our TL-HMM.\ We show the updated latent
state transition diagram in Figure~\ref{fig:trans-perfect}. We can clearly
see that the probability of state 2 has increased dramatically, while the
probability of states 0 and 3 has decreased. State 1 had its probability
increase, but only very slightly.

>>>>

\section{Experiment Results}
As an analysis tool, the TL-HMM model provides with us the following two
patterns to characterize student behavior:
\begin{enumerate*}[label=(\arabic*)]
  \item the latent state representations, and
  \item the latent state transitions.
\end{enumerate*}
Thus to evaluate the proposed model, we conduct experiments to
qualitatively analyze both types of patterns discovered from empirical MOOC
log data.

@@

\begin{figure*}
  \centering
  \begin{tabular}{cccc}
    \textbf{State 0 (Wiki)} & \textbf{State 1 (Lecture)}
    & \textbf{State 2 (Thread View)} & \textbf{State 3 (Lecture/Wiki)}\\\hline
    \includegraphics[width=0.22\textwidth]{figures/text-2state/state0.png}
    &
    \includegraphics[width=0.22\textwidth]{figures/text-2state/state1.png}\\
    \includegraphics[width=0.22\textwidth]{figures/text-3state/state0.png}
    &
    \includegraphics[width=0.22\textwidth]{figures/text-3state/state1.png}
    &
    \includegraphics[width=0.22\textwidth]{figures/text-3state/state2.png}\\
    \includegraphics[width=0.22\textwidth]{figures/text-4state/state0.png}
    &
    \includegraphics[width=0.22\textwidth]{figures/text-4state/state1.png}
    &
    \includegraphics[width=0.22\textwidth]{figures/text-4state/state2.png}
    &
    \includegraphics[width=0.22\textwidth]{figures/text-4state/state3.png}
  \end{tabular}
  \caption{The evolution of states for increasing $K$ for the
  \protect\textretrieval{} MOOC.\ State ``names'' indicate the highest
  probability action(s) within the state.} % I don't know why \protect is needed
  \label{fig:text-state-evolution}
\end{figure*}

@@

\begin{figure*}
  \centering
  \begin{tabular}{cccc}
    \textbf{State 0 (Wiki)}
    & \textbf{State 1 (Lecture/Wiki)}
    & \textbf{State 2 (Thread List)}
    & \textbf{State 3 (Thread View)}\\\hline
    \includegraphics[width=0.22\textwidth]{figures/sustain-2state/state0.png}
    &
    \includegraphics[width=0.22\textwidth]{figures/sustain-2state/state1.png}\\
    \includegraphics[width=0.22\textwidth]{figures/sustain-3state/state0.png}
    &
    \includegraphics[width=0.22\textwidth]{figures/sustain-3state/state2.png}
    &
    \includegraphics[width=0.22\textwidth]{figures/sustain-3state/state1.png}\\
    \includegraphics[width=0.22\textwidth]{figures/sustain-4state/state0.png}
    &
    \includegraphics[width=0.22\textwidth]{figures/sustain-4state/state3.png}
    &
    \includegraphics[width=0.22\textwidth]{figures/sustain-4state/state1.png}
    &
    \includegraphics[width=0.22\textwidth]{figures/sustain-4state/state2.png}
  \end{tabular}
  \caption{The evolution of states for increasing $K$ for the
  \protect\sustain{} MOOC.\ State ``names'' indicate the highest
  probability action(s) within a state.} % I don't know why \protect is needed
  \label{fig:sustain-state-evolution}
\end{figure*}

We observe similar behavior in Figure~\ref{fig:sustain-state-evolution}
as we increase $K$ when fitting our TL-HMM on the \sustain{} MOOC.\ Again,
in the transition between $K=2$ and $K=3$ we discover forum behavior
patterns as a latent state. However, in the transition between $K=3$ to
$K=4$ we instead see a refining of that discovered forum state, where
state 3 captures asymmetric transition probabilities between ``forum thread
view'' and ``forum thread list''. These states could be seen as redundant,
in which case a  setting of $K=3$ may be more appropriate for the
\sustain{} dataset.

@@

Thus, we might expect to see students that perform well in the course
preferring states 1 and 2 over states 0 and 3. To verify this, we took the
model we learned on the full training data and retrofit it to training data
consisting only of sequences produced by students in \textretrieval{} that
had perfect marks. To prevent the latent state meanings from drifting, we
forced the model parameters associated with their Markov model
representations to be fixed, in effect only learning initial and transition
probabilities for the top layer of our TL-HMM.\ We show the updated latent
state transition diagram in Figure~\ref{fig:trans-perfect}. We can clearly
see that the probability of state 2 has increased dramatically, consistent
with previous observations of the positive correlation between forum activity and
grades~\cite{Huang:2014:LAS}, while the probability of states 0 and 3 has
decreased. State 1 had its probability increase, but only very slightly.
